---
title: 'Akanksha Caitlin HW1 '
output:
  html_document: default
  html_notebook: default
---

###Testing the sensitivity of mean, variance and skew to outliers.

We used the rainfall dataset from the class github site to test sensitivity of mean, variance and skewness.
```{r readdata}
rainfall<- read.csv("https://raw.githubusercontent.com/lecondon/EnvDataAnalysis_17/master/R_Sessions/rainfall.csv")

#this is the whole precipitation dataset
precip.all<- rainfall$precip
```

To identify outliers we used a box plot, which calculates outlying values outside 1.5 times the interquartile range.
```{r boxplot}
boxplot(precip.all, horizontal = T)
bpdata <- boxplot(precip.all) 

#this is the boxplot data
bpdata

#these are the outlier values
outliers<- bpdata$out

#removed outliers from the whole precipitation dataset
which(precip.all %in% outliers)
index<- which(precip.all %in% outliers)
precip.clean<- precip.all[-index]

#this is the cleaned dataset without outliers
head(precip.clean)
```
All the outliers in this precipitation dataset are at the high end.

####Running the mean for all the precipitation data and mean for data without outliers.
```{r mean}
meanprecip.all<-mean(precip.all)
meanprecip.clean<- mean(precip.clean)

#create a histogram with kernel density smoother and show each mean
d<- density(precip.all)
{hist(precip.all,freq=F,breaks=seq(0,8000,100), ylim=c(0,6e-4), plot=T, xlab="Annual Precip [mm]", ylab="Probability Density", main="Annual Precipitation") 
lines(d$x, d$y, col='red')
abline(v=meanprecip.all,col="blue")
abline(v=meanprecip.clean,col="green")}
```
The mean decreased slightly when the outliers were removed.

####Running the variance for all the precipitation data and variance for data without outliers
```{r variance}
varprecip.all<-var(precip.all)
varprecip.clean<- var(precip.clean)

#calculate percent change
(varprecip.all-varprecip.clean)/varprecip.all
```
There is a 14% decrease in variance between the whole dataset and the dataset without outliers.
 
####Running skew for the whole dataset and the dataset without outliers.
```{r}
library(moments)
skewprecip.all<- skewness(precip.all)
skewprecip.clean<- skewness(precip.clean)

#show plot of cleaned dataset to see change in skew
d<- density(precip.clean)
{hist(precip.clean,freq=F,breaks=seq(0,8000,100), ylim=c(0,6e-4), plot=T, xlab="Annual Precip [mm]", ylab="Probability Density", main="Annual Precipitation") 
lines(d$x, d$y, col='red')}
```
The whole precipitation dataset has a positve skew because there are extreme values on the high end, so it is being pulled towards right.
The dataset without outliers has a negative skew beacuse there lot of small rain events, so it is being pulled towards left.
